# RNA analysis with DESeq2

```{r message=FALSE}
library(dplyr)
library(dbplyr)
library(readr)
library(DESeq2)
library(pheatmap)
library(rje)
library(ashr)
library(ggplot2)
library(pcaExplorer)
library(tibble)
library(RColorBrewer)
library(AnnotationDbi)
library(AnnotationHub)
library(org.Hs.eg.db)
library(gplots)
library(gage)
library(ensembldb)
library(clusterProfiler)
library(magrittr)
library(EnrichmentBrowser)
library(R.utils)
library(fgsea)
library(data.table)
library(WebGestaltR)
library(UpSetR)
library(gridExtra)
```

## 1. Read count table

```{r}
countdata_NB <- read.delim("count_matrices/countmatrix_NB.csv")
countdata_newIMR <- read.delim("count_matrices/countmatrix_newIMR.csv")
countdata_sensSHY <- read.delim("count_matrices/countmatrix_sensSHY.csv")
countdata_MM <- read.delim("count_matrices/countmatrix_MM.csv")
```


```{r}
countdata_all <- cbind(subset(countdata_NB, select = -c(gene_symbol, Ensembl_ID, length)), subset(countdata_newIMR, select = -c(gene_symbol, Ensembl_ID, length)), subset(countdata_MM, select = -c(gene_symbol, Ensembl_ID, length)), subset(countdata_sensSHY, select = -c(gene_symbol, Ensembl_ID, length)))

rownames(countdata_sensSHY) <- countdata_sensSHY$Ensembl_ID

countdata_sensSHY <- subset(countdata_sensSHY, select = -c(gene_symbol, Ensembl_ID, length))

rownames(countdata_all) <- countdata_NB$Ensembl_ID
```

## 2. Read sample table

```{r}
sampleTable_exp <- read_csv("ont_metadata.csv") %>%
  mutate("Group" = factor(Group)) %>%
  mutate("gender" = factor(gender)) %>%
  mutate("tumor_type" = factor(tumor_type))

head(sampleTable_exp)
```

# Full DESeq data object

```{r}
new_ids <- sampleTable_exp$sample_id_adj[match(colnames(countdata_all), sampleTable_exp$run_id_rnaseq)]
colnames(countdata_all) <- new_ids
head(countdata_all)
```

```{r}
new_ids <- sampleTable_exp$sample_id_adj[match(colnames(countdata_sensSHY), sampleTable_exp$run_id_rnaseq)]
colnames(countdata_sensSHY) <- new_ids
head(countdata_sensSHY)
```

```{r}
countdata_filt <- countdata_all %>%
  dplyr::select(-c("SHY14nov_R3_2", "IMR15dec_R4", "IMR12dec_R3"))

sampleTable_exp_filt = sampleTable_exp[sampleTable_exp$sample_id_adj %in% colnames(countdata_filt), ]
rownames(sampleTable_exp_filt) <- sampleTable_exp_filt$sample_id_adj

# Next, we need to reorder the rows of sampleTable_exp_filt based on the order of the columns in countdata_filt
sampleTable_exp_filt_reordered <- sampleTable_exp_filt[match(colnames(countdata_filt), rownames(sampleTable_exp_filt)), ]

rownames(sampleTable_exp_filt_reordered) <- sampleTable_exp_filt_reordered$sample_id_adj

# Now, we can create the DESeqDataSet using the reordered sample table
dds_full <- DESeqDataSetFromMatrix(countdata_filt, sampleTable_exp_filt_reordered, design = ~ Group)
dds_full$Group <- relevel(dds_full$Group, "S")
dds_full <- estimateSizeFactors(dds_full)

dds_MM <- dds_full[, colData(dds_full)$tumor_type == "MM"]
dds_MM$Group <- factor(droplevels(dds_MM$Group))
dds_NB <- dds_full[, colData(dds_full)$tumor_type == "NB"]

design(dds_full) <- ~ tumor_type + Group
```

```{r}
sampleTable_exp_filt = sampleTable_exp[sampleTable_exp$sample_id_adj %in% colnames(countdata_sensSHY), ]
rownames(sampleTable_exp_filt) <- sampleTable_exp_filt$sample_id_adj

# Next, we need to reorder the rows of sampleTable_exp_filt based on the order of the columns in countdata_filt
sampleTable_exp_filt_reordered <- sampleTable_exp_filt[match(colnames(countdata_sensSHY), rownames(sampleTable_exp_filt)), ]

rownames(sampleTable_exp_filt_reordered) <- sampleTable_exp_filt_reordered$sample_id_adj

dds_sensSHY <- DESeqDataSetFromMatrix(countdata_sensSHY, sampleTable_exp_filt_reordered, design = ~ Group)

dds_sensSHY$Group <- relevel(dds_sensSHY$Group, "Ss")
dds_sensSHY <- estimateSizeFactors(dds_sensSHY)
```


# Sample distances

Calculate a *distance matrix*, quantifying dissimilarity of the samples

```{r}
sample_distance_heatmap <- function(dds, dir, affix) {
  #Used different transformations
  distmat <- as.matrix(dist(t(log2(1 + counts(dds)))))
  distmat_norm <- as.matrix(dist(t(log2(1 + counts(dds, normalized=TRUE)))))
  distmat_vst <- as.matrix(dist(t(assay(vst(dds))))) # Can only be performed if the DE is already performed

  jpeg(paste0(dir, "/Distance_matrix_as_heatmap_normalized_", affix, ".jpeg"), width = 750, height = 750)
  pheatmap(distmat_norm)
  dev.off()
  
  jpeg(paste0(dir, "/Distance_matrix_as_heatmap_vst_", affix, ".jpeg"), width = 750, height = 750)
  pheatmap(distmat_vst)
  dev.off()
  
  # Switch off clustering
  
  jpeg(paste0(dir, "/Distance_matrix_noclus_as_heatmap_normalized_", affix, ".jpeg"), width = 750, height = 750)
  pheatmap(distmat_norm,cluster_rows=FALSE, cluster_cols=FALSE)
  dev.off()
  
  jpeg(paste0(dir, "/Distance_matrix_noclus_as_heatmap_vst_", affix, ".jpeg"), width = 750, height = 750)
  pheatmap(distmat_vst,cluster_rows=FALSE, cluster_cols=FALSE)
  dev.off()
}
```

Visualize the distance matrix as heatmap

```{r fig.width=12, fig.height=12}
affixes <- c("full", "NB", "MM", "sensSHY")
dds_list <- list(dds_full, dds_NB, dds_MM, dds_sensSHY)
dir <- "outputs/DataExploration"
dir.create(dir)

for (i in seq_along(dds_list)) {
  sample_distance_heatmap(dds_list[[i]], dir, affixes[i])
}
```

# PCA plot

```{r}
dir <- "outputs/DataExploration"
dir.create(dir)

rld_full <- vst(dds_full)
rld_NB <- vst(dds_NB)
rld_MM <- vst(dds_MM)
rld_sensSHY <- vst(dds_sensSHY)

## Top 500

png(paste0(dir, "/PCA_full_ntop500.png"), width = 900, height = 900)
pca_full_ntop500 <- pcaplot(rld_full, title = "PCA plot - top 500 variable genes - full", intgroup=c("tumor_type", "Group"), ellipse = TRUE, text_labels = TRUE)
pca_full_ntop500
dev.off()

png(paste0(dir, "/PCA_NB_ntop500.png"), width = 900, height = 900)
pca_NB_ntop500 <- pcaplot(rld_NB, title = "PCA plot - top 500 variable genes - NB", intgroup=c("tumor_type", "Group"), ellipse = TRUE, text_labels = TRUE)
pca_NB_ntop500
dev.off()

png(paste0(dir, "/PCA_MM_ntop500.png"), width = 900, height = 900)
pca_MM_ntop500 <- pcaplot(rld_MM, title = "PCA plot - top 500 variable genes - MM", intgroup=c("tumor_type", "Group"), ellipse = TRUE, text_labels = TRUE)
pca_MM_ntop500
dev.off()

png(paste0(dir, "/PCA_sensSHY_ntop500.png"), width = 900, height = 900)
pca_sensSHY_ntop500 <- pcaplot(rld_sensSHY, title = "PCA plot - top 500 variable genes - sensSHY", intgroup=c("tumor_type", "Group"), ellipse = TRUE, text_labels = TRUE)
pca_sensSHY_ntop500
dev.off()

png(paste0(dir, "/PCA_combined_plots.png"), width = 900, height = 900)
pca_combined_ntop500 <- grid.arrange(pca_full_ntop500, pca_NB_ntop500, pca_MM_ntop500, pca_sensSHY_ntop500, ncol=2)
pca_combined_ntop500
dev.off()

## Top 10000

png(paste0(dir, "/PCA_full_ntop10000.png"), width = 900, height = 900)
pcaplot(rld_full, intgroup=c("tumor_type", "Group"), ellipse = TRUE, text_labels = TRUE, ntop = 10000)
dev.off()

png(paste0(dir, "/PCA_NB_ntop10000.png"), width = 900, height = 900)
pcaplot(rld_NB, intgroup=c("tumor_type", "Group"), ellipse = TRUE, text_labels = TRUE, ntop = 10000)
dev.off()

png(paste0(dir, "/PCA_MM_ntop10000.png"), width = 900, height = 900)
pcaplot(rld_MM, intgroup=c("tumor_type", "Group"), ellipse = TRUE, text_labels = TRUE, ntop = 10000)
dev.off()

png(paste0(dir, "/PCA_sensSHY_ntop10000.png"), width = 900, height = 900)
pcaplot(rld_sensSHY, intgroup=c("tumor_type", "Group"), ellipse = TRUE, text_labels = TRUE, ntop = 10000)
dev.off()
```


# Two-group comparison

For our setup, we want to compare R and S

## 1. Size factors and metadata

```{r}
# Calculate sizeFactors (should be 1)

for (i in seq_along(dds_list)) {
  dds <- dds_list[[i]]
  print(affixes[i])
  print(sizeFactors(dds))
}
```


## 2. Set design

```{r}
dir.create("outputs/DifferentialAnalysis")

dds_deseq_list = list()

for (i in seq_along(dds_list)) {
  dds <- dds_list[[i]]
  affix <- affixes[[i]]
  print(affix)

  dds_deseq <- DESeq(dds)
  
  if (affix == "full" | affix == "NB") {
    contrasts <- list(c("Group", "R", "S"), c("Group", "R", "Ss"), c("Group", "S", "Ss"))
  } else if (affix == "sensSHY") {
    contrasts <- list(c("Group", "R", "Ss"))
  } else {
    contrasts <- list(c("Group", "R", "S"))
  }
  
  for (contrast in contrasts) {
    print(paste(contrast, collapse = "_"))
    res <- results(dds_deseq, alpha = 0.05, contrast = contrast)
    summary(res)
    jpeg(paste0("outputs/DifferentialAnalysis/dispersion_plot_", affix, "_", paste(contrast, collapse = "_"), ".jpeg"), 
         width = 1500, height = 1000)
    plotDispEsts(dds_deseq)
    dev.off()
    
    dds_deseq_list <- c(dds_deseq_list, dds_deseq)
  }
}
```
```{r}
contrasts <- list(c("Group", "R", "S"), c("Group", "R", "Ss"), c("Group", "S", "Ss"), c("Group", "R", "S"), c("Group", "R", "Ss"), c("Group", "S", "Ss"), c("Group", "R", "S"), c("Group", "R", "Ss"))
affixes <- list("full", "full", "full", "NB", "NB", "NB", "MM", "sensSHY")

# Combine the elements from both lists
combined_list <- gsub(" ", "", paste0(affixes, "_", sapply(contrasts, function(x) paste(x[length(x) - 1], "_vs_", x[length(x)], collapse = ""))))

# Print the combined list
print(combined_list)
```


## 3. Get the results

```{r}
res_lfc_list <- list()

for (i in seq_along(dds_deseq_list)) {
  dds <- dds_deseq_list[[i]]
  affix <- affixes[i]
  print(affix)
  contrast = contrasts[[i]]
  print(paste(contrast, collapse = "_"))
   
  resultsNames(dds)
  res <- results(dds, independentFiltering=TRUE, contrast=contrast, alpha = 0.05) #Last level is the base level
  res_all <- results(dds, independentFiltering=TRUE, contrast=contrast)
  
  sink(paste0("outputs/DifferentialAnalysis/results_summary_", affix, "_", paste(contrast, collapse = "_"), ".txt"))
  summary(res)
  sink()
  
  res_lfc <- lfcShrink(dds, contrast = contrast, type = "ashr", alpha = 0.05, res = res)
  res_lfc_all <- lfcShrink(dds, contrast = contrast, type = "ashr", res = res_all)
  
  sink(paste0("outputs/DifferentialAnalysis/resultsLFC_summary_", affix, "_", paste(contrast, collapse = "_"), ".txt"))
  summary(res_lfc)
  sink()
  
  res_lfc_list <- c(res_lfc_list, res_lfc_all)
}
```

## PCA (DGE only)

```{r}
# Create a list to store PCA plots
pca_plots <- vector("list", length = length(dds_deseq_list))

# Perform PCA and generate plots
for (i in seq_along(dds_deseq_list)) {
  dds <- dds_deseq_list[[i]]
  res <- res_lfc_list[[i]] 
  affix <- affixes[i]
  print(affix)
  contrast = contrasts[[i]]
  print(paste(contrast, collapse = "_"))
  
  rld <- vst(dds)
  
  dge_genes <- rownames(res)[which(res$padj < 0.05)]
  
  # Filter expression data matrices to include only DEGs
  rld_dge <- rld[dge_genes,]
  
  # Perform PCA
  pca_result <- prcomp(t(assay(rld_dge)))
  
  # Store PCA result
  pca_plots[[i]] <- list(pca_result = pca_result, affix = affix)
}

# Generate PCA plots
for (i in seq_along(pca_plots)) {
  pca_result <- pca_plots[[i]]$pca_result
  affix <- pca_plots[[i]]$affix
  contrast = contrasts[[i]]
  
  png(paste0(dir, paste0("/PCA_", affix, "_", paste(contrast, collapse = "_"), "_dge.png")), width = 900, height = 900)
  
  plot(pca_result$x[,1], pca_result$x[,2], 
       xlab = paste("PC1 (", round(100 * summary(pca_result)$importance[2,1], 1), "%)", sep=""),
       ylab = paste("PC2 (", round(100 * summary(pca_result)$importance[2,2], 1), "%)", sep=""), 
       main = paste("PCA of", affix, "_", paste(contrast, collapse = "_"), "DEGs"))
  
  # Add labels to points
  text(pca_result$x[,1], pca_result$x[,2], labels = rownames(pca_result$x), cex=0.7, pos=3)
  
  dev.off()
}
```


Other plotting functions
## 5. Plotting: PCA, MA-plot, Volcano, heatmap, etc...

```{r}
# Set generic dds and res, condition, resdata and resdata_annot

###########SECTION TO BE ADJUSTED#######
condition_name <- "Group"
celline <- "human"
###########SECTION TO BE ADJUSTED#######

resdata_list <- list()

for (i in seq_along(dds_deseq_list)) {
  res <- res_lfc_list[[i]]
  dds <- dds_deseq_list[[i]]
  affix <- affixes[i]
  print(affix)
  
  resdata <- merge(as.data.frame(res), as.data.frame(counts(dds, normalized=TRUE)), by="row.names", sort=FALSE)

  names(resdata)[1] <- "Gene"
  rownames(resdata) <- resdata$Gene
  resdata <- resdata[-1]
  head(resdata)
  
  resdata_list <- c(resdata_list, list(resdata))
}
```


```{r}
annotdata <- read.table("mart_export.txt",sep = '\t', header=TRUE)
colnames(annotdata) <- c('ensg','name','type','description')

#Check for non-unique ensg and remove duplicates if necessary
n_occur <- data.frame(table(annotdata$ensg))
annotdata[annotdata$ensg %in% n_occur$Var1[n_occur$Freq == 1],]

# Manually erase from annotation file where necessary

rownames(annotdata) <- annotdata$ensg
annotdata <- annotdata[-1]
head(annotdata)

resdata_annot_list <- list()

for (i in seq_along(resdata_list)) {
  resdata <- resdata_list[[i]]
  affix <- affixes[i]
  print(affix)
  
  resdata_annot <- merge(as.data.frame(resdata),as.data.frame(annotdata),by="row.names",all.x= TRUE)
  rownames(resdata_annot) <- resdata_annot$Row.names
  resdata_annot <- resdata_annot[-1]
  
  resdata_annot_list <- c(resdata_annot_list, list(resdata_annot))
}
```


```{r}
# Regularized log transformation for clustering/heatmaps, etc

rld_list <- list()

for (i in seq_along(dds_deseq_list)) {
  dds <- dds_deseq_list[[i]]
  affix <- affixes[i]
  print(affix)
  contrast = contrasts[[i]]
  print(contrast)
  
  rld <- vst(dds)
  head(assay(rld))
  
  png(paste0("outputs/DifferentialAnalysis/rld_hist_", affix, "_", paste(contrast, collapse = "_"), ".png"))
  hist(assay(rld))
  dev.off()
  
  rld_list <- c(rld_list, rld)
}
```


```{r}
condition_list <- list()

for (i in seq_along(dds_deseq_list)) {
  dds <- dds_deseq_list[[i]]
  rld <- rld_list[[i]]
  affix <- affixes[i]
  print(affix)
  contrast = contrasts[[i]]
  print(contrast)
  
  condition <- colData(dds)$Group
  
  mycols <- brewer.pal(8, "Dark2")[1:length(unique(condition))]

  # Sample distance heatmap
  
  sampleDists <- as.matrix(dist(t(assay(rld))))
  
  png(paste0("outputs/DifferentialAnalysis/heatmap_sampdist_", affix, "_", paste(contrast, collapse = "_"), ".png"))
  heatmap.2(as.matrix(sampleDists), key=F, trace="none",
            col=colorpanel(100, "black", "white"),
            ColSideColors=mycols[condition], RowSideColors=mycols[condition],
            margin=c(20, 20), main="Sample Distance Matrix")
  dev.off()
  
  condition_list <- list(condition_list, condition)
}
```


```{r}
for (i in seq_along(resdata_annot_list)) {
  resdata_annot <- resdata_annot_list[[i]]
  res <- res_lfc_list[[i]]
  affix <- affixes[i]
  print(affix)
  contrast = paste(contrasts[[i]], collapse = "_")
  print(contrast)
  
  DE_file_name <- paste("outputs/DifferentialAnalysis/DE_results_",celline,"_",contrast,"_", affix,".csv", sep="" )

  ## Order by adjusted p-value
  resdata_annot <- resdata_annot[order(resdata_annot$padj), ]
  
  ## Write results
  write.csv(resdata_annot, DE_file_name)
  
  ## Examine plot of p-values
  png(paste("outputs/DifferentialAnalysis/DE_pvalplot_",celline,"_",contrast,"_", affix, ".png", sep=""), 1500, 1000, pointsize=20)
  hist(resdata_annot$pvalue, breaks=50, col="grey")
  dev.off()
  
  ## MA-plot
  
  maplot <- function (res, thresh=0.05, labelsig=TRUE, textcx=1, ...) {
  with(res, plot(baseMean, log2FoldChange, pch=20, cex=.5, log="x", ...))
  with(subset(res, padj<thresh), points(baseMean, log2FoldChange, col="red", pch=20, cex=0.5))
  if (labelsig) {
    require(calibrate)
    }
  }
  
  png(paste("outputs/DifferentialAnalysis/DE_MAplot_",celline,"_",contrast,"_", affix, ".png", sep=""), 1500, 1000, pointsize=20)
  maplot(resdata_annot, main="MA Plot", ylim=c(-10,10))
  dev.off()
  
  resdata_annot_notnull <- resdata_annot %>% dplyr::filter(padj !=0)
  minpvaladj <- min(resdata_annot_notnull$padj, na.rm = TRUE)
  
  ## Volcano plot with "significant" genes labeled
  volcanoplot <- function (res, lfcthresh=1, sigthresh=0.05, main="Volcano Plot", legendpos="bottomleft", labelsig=TRUE, textcx=1, ...) {
    with(res, plot(log2FoldChange, -log10(pvalue+minpvaladj), pch=20, main=main, ...))
    with(subset(res, padj<sigthresh ), points(log2FoldChange, -log10(pvalue+minpvaladj), pch=20, col="red", ...))
    with(subset(res, abs(log2FoldChange)>lfcthresh), points(log2FoldChange, -log10(pvalue+minpvaladj), pch=20, col="orange", ...))
    with(subset(res, padj<sigthresh & abs(log2FoldChange)>lfcthresh), points(log2FoldChange, -log10(pvalue+minpvaladj), pch=20, col="green", ...))
    if (labelsig) {
      require(calibrate)
    }
    legend(legendpos, xjust=1, yjust=1, legend=c(paste("FDR<",sigthresh,sep=""), paste("|LogFC|>",lfcthresh,sep=""), "both"), pch=20, col=c("red","orange","green"))
  }
  png(paste("outputs/DifferentialAnalysis/DE_volcano_",celline,"_",contrast,"_", affix, ".png", sep=""), 1200, 1000, pointsize=20)
  volcanoplot(resdata_annot, lfcthresh=1, sigthresh=0.05, textcx=.8, xlim=c(-10, 10), ylim=c(0,10))
  dev.off()
}
```


# Heatmap of top expression results

```{r}

###########SECTION TO BE ADJUSTED#######
condition_name <- "Group"
celline <- "human"
###########SECTION TO BE ADJUSTED#######

for (i in seq_along(dds_deseq_list)) {
  dds <- dds_deseq_list[[i]]
  res <- res_lfc_list[[i]]
  affix <- affixes[i]
  print(affix)
  contrast = paste(contrasts[[i]], collapse = "_")
  print(contrast)
  
  vsd <- vst(dds)
  vsd_assay <- assay(vsd)
  vsd_assay <- as.data.frame(vsd_assay)
  vsd_assay$Gene <- rownames(vsd_assay)
  
  ## Merge with normalized count data
  resdata <- merge(as.data.frame(res), vsd_assay, by="row.names", sort=FALSE)
  names(resdata)[1] <- "Gene"
  rownames(resdata) <- resdata$Gene
  resdata <- resdata[-1]
  
  #Merge with annotation (genesymbol and geneType)
  resdata_annot <- merge(as.data.frame(resdata),as.data.frame(annotdata),by="row.names",all.x= TRUE)
  rownames(resdata_annot) <- resdata_annot$Row.names
  resdata_annot <- resdata_annot[-1]
  
  # Selection on padj <= 0.05
  res$padj <- ifelse(is.na(res$padj), 1, res$padj)
  sigGenes <- rownames(res[res$padj <= .05,])
  
  resdata_annot <- resdata_annot[resdata_annot$Gene %in% sigGenes,]
  
  mat = resdata_annot[ head(order(resdata_annot$padj),50), ] # select the top 50 genes with the lowest padj
  rownames(mat) <- paste(mat$Gene, mat$name, sep="_")
  mat$log2FoldChange <- NULL
  mat$lfcSE <- NULL
  mat$stat <- NULL
  mat$pvalue <- NULL
  mat$baseMean <- NULL
  mat$padj <- NULL
  mat$Gene <- NULL
  mat$name <- NULL
  mat$type <- NULL
  mat$description <- NULL
  
  mat = mat - rowMeans(mat) # Subtract the row means from each value
  # Optional, but to make the plot nicer:
  df = as.data.frame(colData(vsd)[,c("Group")]) # Create a dataframe with a column of the conditions
  colnames(df) = "Group" # Rename the column header
  rownames(df) = colnames(mat) # add rownames
  # and plot the actual heatmap
  png(paste("outputs/DifferentialAnalysis/genes_heatmap_",celline,"_",contrast,"_", affix, ".png", sep=""), 1200, 1000, pointsize=20)
  pheatmap(mat, annotation_col=df)
  dev.off()
}
```

```{r}
for (i in seq_along(dds_deseq_list)) {
  dds <- dds_deseq_list[[i]]
  res <- res_lfc_list[[i]]
  affix <- affixes[i]
  print(affix)
  
  rld <- vst(dds)
  
  # make the lists
  upgenes <- rownames(head(res[ order( res$log2FoldChange ), ], n=20))
  downgenes <- rownames(head(res[ order( -res$log2FoldChange ), ], n=20))

  # this gives us the rows we want
  rows <- match(upgenes, row.names(rld))
  mat <- assay(rld)[rows,]
  mat <- mat - rowMeans(mat)

  # the labels are hard coded at the moment :(
  df <- as.data.frame(colData(rld)[c(condition_name)])
  png(paste("outputs/DifferentialAnalysis/top20UP_",celline,"_",contrast,"_", affix, ".png", sep=""))
  pheatmap(mat, fontsize=5, annotation_col=df, main=paste(affix,"top 20 up genes"))
  dev.off()

  # this gives us the rows we want
  rows <- match(downgenes, row.names(rld))
  mat <- assay(rld)[rows,]
  mat <- mat - rowMeans(mat)

  df <- as.data.frame(colData(rld)[c(condition_name)])
  png(paste("outputs/DifferentialAnalysis/top20DOWN_",celline,"_",contrast,"_", affix, ".png", sep=""))
  pheatmap(mat, fontsize=5, annotation_col=df, main=paste(affix,"top 20 down genes"))
  dev.off()
}
```

```{r}
# Example data (replace this with your own data)
gene_sets_up <- list()
gene_sets_down <- list()

for (i in seq_along(resdata_annot_list)) {
  title = combined_list[[i]]
  resdata_annot <- resdata_annot_list[[i]] %>%
    filter(padj < 0.05) %>%
    filter(log2FoldChange > 0)
  gene_sets_up[[title]] <- unique(na.omit(resdata_annot$name))
}

for (i in seq_along(resdata_annot_list)) {
  title = combined_list[[i]]
  resdata_annot <- resdata_annot_list[[i]] %>%
    filter(padj < 0.05) %>%
    filter(log2FoldChange < 0)
  gene_sets_down[[title]] <- unique(na.omit(resdata_annot$name))
}

whitelist = read.csv("complete_gene_list_no_duplicates.txt")

gene_sets_up[["whitelist"]] = whitelist[, 2]
gene_sets_down[["whitelist"]] = whitelist[, 2]

generate_upset_plot <- function(gene_sets, ylabel, file){
  # Create UpSet plot
  upset_data <- fromList(gene_sets)
  upset_plot <- upset(upset_data, 
                      sets = names(gene_sets), 
                      keep.order = TRUE, 
                      mainbar.y.label = ylabel,
                      nintersects = 20,
                      text.scale = 3)
  
  print(upset_plot)
  
  # Save the plot as a PNG file with increased resolution
  png(file, width = 1900, height = 1200, pointsize = 5)
  print(upset_plot)
  dev.off()
}

generate_upset_plot(gene_sets_up, "Gene Symbol Intersection Size - Upregulated", "outputs/DifferentialAnalysis/upset_plot_up.png")
generate_upset_plot(gene_sets_down, "Gene Symbol Intersection Size - Downregulated","outputs/DifferentialAnalysis/upset_plot_down.png")
```


