# RNA analysis with DESeq2

```{r message=FALSE}
library(dplyr)
library(dbplyr)
library(readr)
library(DESeq2)
library(pheatmap)
library(rje)
library(ashr)
library(ggplot2)
library(pcaExplorer)
library(tibble)
library(RColorBrewer)
library(AnnotationDbi)
library(AnnotationHub)
library(org.Hs.eg.db)
library(gplots)
library(gage)
library(ensembldb)
library(clusterProfiler)
library(magrittr)
library(EnrichmentBrowser)
library('R.utils')
library(fgsea)
library(data.table)
library(WebGestaltR)
```

## 1. Read count table

```{r}
#dir <- "/data/lvisser/rnaseq_ferroptomics/"

countdata_fc_raw <- read.delim("rnaseq_ferroptomics/outputs/count_matrix/countmatrix.csv")
countdataTPM_fc_raw <- read.delim("rnaseq_ferroptomics/outputs/count_matrix/countmatrix_tpm.csv")

countdata_fc_full <- countdata_fc_raw %>%
  dplyr::select(-c("gene_symbol", "Ensembl_ID", "length"))


rownames(countdata_fc_full) <- countdata_fc_raw$Ensembl_ID

print(colSums(is.na(countdata_fc_full)))
print(dim(countdata_fc_full))
print(colnames(countdata_fc_full))
```

## 2. Read sample table

```{r}
sampleTable_exp <- read_csv("rnaseq_ferroptomics/MT_ferroptosis_metadata.csv") %>%
  mutate("Group" = factor(Group))

head(sampleTable_exp)
print(dim(sampleTable_exp))
```

# Full DESeq data object

```{r}
new_ids <- sampleTable_exp$sample_id_adj[match(colnames(countdata_fc_full), sampleTable_exp$run_id)]
colnames(countdata_fc_full) <- new_ids
countdata_fc_full <- countdata_fc_full[, sampleTable_exp$sample_id_adj]
head(countdata_fc_full)

countdata_fc_filt1 <- countdata_fc_full %>%
  dplyr::select(-c("SHY14nov_R3_2"))

countdata_fc_filt2 <- countdata_fc_filt1 %>%
  dplyr::select(-c("IMR15dec_R4", "IMR12dec_R3"))

countdata_fc_filt3 <- countdata_fc_filt2 %>%
  dplyr::select(-c("IMR21nov_R2", "IMR14nov_R1"))

dds_full <- DESeqDataSetFromMatrix(countdata_fc_full, sampleTable_exp, ~1)
dds_full$Group <- relevel(dds_full$Group, "S")
dds_full <- estimateSizeFactors(dds_full)

dds_filt1 <- dds_full[, !(colData(dds_full)$sample_id_adj %in% c("SHY14nov_R3_2"))]
dds_filt2 <- dds_filt1[, !(colData(dds_filt1)$sample_id_adj %in% c("IMR15dec_R4", "IMR12dec_R3"))]
dds_filt3 <- dds_filt2[, !(colData(dds_filt2)$sample_id_adj %in% c("IMR21nov_R2", "IMR14nov_R1"))]
```

# Sample distances

Calculate a *distance matrix*, quantifying dissimilarity of the samples

```{r}
sample_distance_heatmap <- function(dds, dir, affix) {
  #Used different transformations
  distmat <- as.matrix(dist(t(log2(1 + counts(dds)))))
  distmat_norm <- as.matrix(dist(t(log2(1 + counts(dds, normalized=TRUE)))))
  distmat_vst <- as.matrix(dist(t(assay(vst(dds))))) # Can only be performed if the DE is already performed

  jpeg(paste0(dir, "/Distance_matrix_as_heatmap_normalized_", affix, ".jpeg"), width = 750, height = 750)
  pheatmap(distmat_norm)
  dev.off()
  
  jpeg(paste0(dir, "/Distance_matrix_as_heatmap_vst_", affix, ".jpeg"), width = 750, height = 750)
  pheatmap(distmat_vst)
  dev.off()
  
  # Switch off clustering
  
  jpeg(paste0(dir, "/Distance_matrix_noclus_as_heatmap_normalized_", affix, ".jpeg"), width = 750, height = 750)
  pheatmap(distmat_norm,cluster_rows=FALSE, cluster_cols=FALSE)
  dev.off()
  
  jpeg(paste0(dir, "/Distance_matrix_noclus_as_heatmap_vst_", affix, ".jpeg"), width = 750, height = 750)
  pheatmap(distmat_vst,cluster_rows=FALSE, cluster_cols=FALSE)
  dev.off()
}
```

Visualize the distance matrix as heatmap

```{r fig.width=12, fig.height=12}
affixes <- c("full", "filt1", "filt2", "filt3")
dds_list <- list(dds_full, dds_filt1, dds_filt2, dds_filt3)
dir <- "rnaseq_ferroptomics/DataExploration"
dir.create(dir)

for (i in seq_along(dds_list)) {
  sample_distance_heatmap(dds_list[[i]], dir, affixes[i])
}
```

# MATRIX OF SCATTERPLOTS

```{r}

# panel.smooth function is built in.
# panel.cor puts correlation in upper panels, size proportional to correlation
panel.cor <- function(x, y, digits=3, prefix="", cex.cor, ...)
{
    usr <- par("usr"); on.exit(par(usr))
    par(usr = c(0, 1, 0, 1))
    r <- abs(cor(x, y,use=c("complete.obs")))
    txt <- format(c(r, 0.123456789), digits=digits)[1]
    txt <- paste(prefix, txt, sep="")
    if(missing(cex.cor)) cex.cor <- 0.8/strwidth(txt)
    text(0.5, 0.5, txt, cex = cex.cor * r)
}
# Plot: add loess smoother in lower and correlation in upper
#Add +columnA+columnB to add matrix dimensions

# jpeg("rnaseq_ferroptomics/DataExploration/RNAseq_scatterplot_matrix_bw.jpeg", width = 1500, height = 1500)
# pairs(log10(countdata_fc+1), lower.panel=panel.smooth, upper.panel=panel.cor,main="Scatterplot Matrix RNAseq")
# dev.off()
# 
# jpeg("rnaseq_ferroptomics/DataExploration/RNAseq_Scatterplot_matrix_color.jpeg", width = 750, height = 750)
# pairs(log10(countdata_fc+1), 
#      lower.panel = function(...) smoothScatter(..., colramp = colorRampPalette(c(rev(brewer.pal(5,"RdYlBu")))), add = TRUE, nrpoints=0), upper.panel=panel.cor, pch=20, main="Scatterplot Matrix RNAseq")
# dev.off()

```

# PCA plot

```{r}
dir <- "rnaseq_ferroptomics/DataExploration"
dir.create(dir)

rld_full <- vst(dds_full)
rld_filt1 <- vst(dds_filt1)
rld_filt2 <- vst(dds_filt2)
rld_filt3 <- vst(dds_filt3)

png(paste0(dir, "/PCA_full.png"), width = 900, height = 900)
pcaplot(rld_full, intgroup=c("Group"), ellipse = FALSE)
dev.off()

png(paste0(dir, "/PCA_filt1.png"), width = 900, height = 900)
pcaplot(rld_filt1, intgroup=c("Group"), ellipse = FALSE)
dev.off()

png(paste0(dir, "/PCA_filt2.png"), width = 900, height = 900)
pcaplot(rld_filt2, intgroup=c("Group"), ellipse = FALSE)
dev.off()

png(paste0(dir, "/PCA_filt3.png"), width = 900, height = 900)
pcaplot(rld_filt3, intgroup=c("Group"), ellipse = FALSE)
dev.off()
```


# Two-group comparison

For our setup, we want to compare R and S

## 1. Size factors and metadata

```{r}

# Calculate sizeFactors (should be 1)

for (i in seq_along(dds_list)) {
  dds <- dds_list[[i]]
  print(affixes[i])
  print(sizeFactors(dds))
  # print(colData(dds))
}
```

## 2. Set design

Comparing two types of samples, which differ by a single covariate (here: `Group`), ("a two-group comparison") is straight-forward.

Set the comparison type, i.e. indicate which column in the colData table contains the group information


```{r}
dir.create("rnaseq_ferroptomics/DifferentialAnalysis")

dds_deseq_list = list()

for (i in seq_along(dds_list)) {
  dds <- dds_list[[i]]
  affix <- affixes[[i]]
  print(affix)
  
  design(dds) <- ~ Group

  dds_deseq <- DESeq(dds)
  
  res <- results(dds_deseq, alpha = 0.05)
  summary(res)
  
  jpeg(paste0("rnaseq_ferroptomics/DifferentialAnalysis/dispersion_plot_", affix, ".jpeg"), width = 1500, height = 1000)
  plotDispEsts(dds_deseq)
  dev.off()
  
  dds_deseq_list <- c(dds_deseq_list, dds_deseq)
}
```

## 3. Get the results

```{r}
res_RvsS_lfc_list <- list()

for (i in seq_along(dds_deseq_list)) {
  dds <- dds_deseq_list[[i]]
  affix <- affixes[i]
  print(affix)
  
  resultsNames(dds)
  res_RvsS <- results(dds, independentFiltering=TRUE, contrast=c("Group","R","S"), alpha = 0.05) #Last level is the base level
  summary(res_RvsS)
  
  res_RvsS_lfc <- lfcShrink(dds, contrast = c("Group", "R", "S"), type = "ashr", lfcThreshold = 0, res = res_RvsS)
  
  summary(res_RvsS_lfc)
  
  res_RvsS_lfc_list <- c(res_RvsS_lfc_list, res_RvsS_lfc)
}

###CONSTRUCT THE SUMMARY TABLE OF ALL PADJ <= 0.05

contrast <- "RvsS"
```

Other plotting functions
## 5. Plotting: PCA, MA-plot, Volcano, heatmap, etc...

```{r}
# Set generic dds and res, condition, resdata and resdata_annot

###########SECTION TO BE ADJUSTED#######
contrast <- "RvsS"
condition_name <- "Group"
celline <- "human"
###########SECTION TO BE ADJUSTED#######

resdata_list <- list()

for (i in seq_along(dds_deseq_list)) {
  res <- res_RvsS_lfc_list[[i]]
  dds <- dds_deseq_list[[i]]
  affix <- affixes[i]
  print(affix)
  
  resdata <- merge(as.data.frame(res), as.data.frame(counts(dds, normalized=TRUE)), by="row.names", sort=FALSE)

  names(resdata)[1] <- "Gene"
  rownames(resdata) <- resdata$Gene
  resdata <- resdata[-1]
  head(resdata)
  
  resdata_list <- c(resdata_list, list(resdata))
}
```


```{r}
annotdata <- read.table("rnaseq_ferroptomics/mart_export.txt",sep = '\t', header=TRUE)
colnames(annotdata) <- c('ensg','name','type','description')

#Check for non-unique ensg and remove duplicates if necessary
n_occur <- data.frame(table(annotdata$ensg))
annotdata[annotdata$ensg %in% n_occur$Var1[n_occur$Freq == 1],]

# Manually erase from annotation file where necessary

rownames(annotdata) <- annotdata$ensg
annotdata <- annotdata[-1]
head(annotdata)

resdata_annot_list <- list()

for (i in seq_along(resdata_list)) {
  resdata <- resdata_list[[i]]
  affix <- affixes[i]
  print(affix)
  
  resdata_annot <- merge(as.data.frame(resdata),as.data.frame(annotdata),by="row.names",all.x= TRUE)
  rownames(resdata_annot) <- resdata_annot$Row.names
  resdata_annot <- resdata_annot[-1]
  
  resdata_annot_list <- c(resdata_annot_list, list(resdata_annot))
}
```


```{r}
# Regularized log transformation for clustering/heatmaps, etc

rld_list <- list()

for (i in seq_along(dds_deseq_list)) {
  dds <- dds_deseq_list[[i]]
  affix <- affixes[i]
  print(affix)
  
  rld <- vst(dds)
  head(assay(rld))
  
  png(paste0("rnaseq_ferroptomics/DifferentialAnalysis/rld_hist_", affix, ".png"))
  hist(assay(rld))
  dev.off()
  
  rld_list <- c(rld_list, rld)
}
```


```{r}
condition_list <- list()

for (i in seq_along(dds_deseq_list)) {
  dds <- dds_deseq_list[[i]]
  rld <- rld_list[[i]]
  affix <- affixes[i]
  print(affix)
  
  condition <- colData(dds)$Group
  
  mycols <- brewer.pal(8, "Dark2")[1:length(unique(condition))]

  # Sample distance heatmap
  
  sampleDists <- as.matrix(dist(t(assay(rld))))
  
  png(paste0("rnaseq_ferroptomics/DifferentialAnalysis/heatmap_sampdist_", affix, ".png"))
  heatmap.2(as.matrix(sampleDists), key=F, trace="none",
            col=colorpanel(100, "black", "white"),
            ColSideColors=mycols[condition], RowSideColors=mycols[condition],
            margin=c(20, 20), main="Sample Distance Matrix")
  dev.off()
  
  condition_list <- list(condition_list, condition)
}
```


```{r}
for (i in seq_along(resdata_annot_list)) {
  resdata_annot <- resdata_annot_list[[i]]
  res <- res_RvsS_lfc_list[[i]]
  affix <- affixes[i]
  print(affix)
  
  DE_file_name <- paste("rnaseq_ferroptomics/DifferentialAnalysis/DE_results_",celline,"_",contrast,"_", affix,".csv", sep="" )

  ## Order by adjusted p-value
  resdata_annot <- resdata_annot[order(resdata_annot$padj), ]
  
  ## Write results
  write.csv(resdata_annot, DE_file_name)
  
  ## Examine plot of p-values
  png(paste("rnaseq_ferroptomics/DifferentialAnalysis/DE_pvalplot_",celline,"_",contrast,"_", affix, ".png", sep=""), 1500, 1000, pointsize=20)
  hist(resdata_annot$pvalue, breaks=50, col="grey")
  dev.off()
  
  ## MA-plot
  
  maplot <- function (res, thresh=0.05, labelsig=TRUE, textcx=1, ...) {
  with(res, plot(baseMean, log2FoldChange, pch=20, cex=.5, log="x", ...))
  with(subset(res, padj<thresh), points(baseMean, log2FoldChange, col="red", pch=20, cex=0.5))
  if (labelsig) {
    require(calibrate)
    }
  }
  
  png(paste("rnaseq_ferroptomics/DifferentialAnalysis/DE_MAplot_",celline,"_",contrast,"_", affix, ".png", sep=""), 1500, 1000, pointsize=20)
  maplot(resdata_annot, main="MA Plot", ylim=c(-10,10))
  dev.off()
  
  resdata_annot_notnull <- resdata_annot %>% dplyr::filter(padj !=0)
  minpvaladj <- min(resdata_annot_notnull$padj, na.rm = TRUE)
  
  ## Volcano plot with "significant" genes labeled
  volcanoplot <- function (res, lfcthresh=2, sigthresh=0.05, main="Volcano Plot", legendpos="bottomleft", labelsig=TRUE, textcx=1, ...) {
    with(res, plot(log2FoldChange, -log10(pvalue+minpvaladj), pch=20, main=main, ...))
    with(subset(res, padj<sigthresh ), points(log2FoldChange, -log10(pvalue+minpvaladj), pch=20, col="red", ...))
    with(subset(res, abs(log2FoldChange)>lfcthresh), points(log2FoldChange, -log10(pvalue+minpvaladj), pch=20, col="orange", ...))
    with(subset(res, padj<sigthresh & abs(log2FoldChange)>lfcthresh), points(log2FoldChange, -log10(pvalue+minpvaladj), pch=20, col="green", ...))
    if (labelsig) {
      require(calibrate)
    }
    legend(legendpos, xjust=1, yjust=1, legend=c(paste("FDR<",sigthresh,sep=""), paste("|LogFC|>",lfcthresh,sep=""), "both"), pch=20, col=c("red","orange","green"))
  }
  png(paste("rnaseq_ferroptomics/DifferentialAnalysis/DE_volcano_",celline,"_",contrast,"_", affix, ".png", sep=""), 1200, 1000, pointsize=20)
  volcanoplot(resdata_annot, lfcthresh=1, sigthresh=0.05, textcx=.8, xlim=c(-10, 10), ylim=c(0,10))
  dev.off()
}
```

# Heatmap of top expression results

```{r}

###########SECTION TO BE ADJUSTED#######
contrast <- "RvsS"
condition_name <- "Group"
celline <- "human"
###########SECTION TO BE ADJUSTED#######

for (i in seq_along(dds_deseq_list)) {
  dds <- dds_deseq_list[[i]]
  res <- res_RvsS_lfc_list[[i]]
  affix <- affixes[i]
  print(affix)
  
  vsd <- vst(dds)
  vsd_assay <- assay(vsd)
  vsd_assay <- as.data.frame(vsd_assay)
  vsd_assay$Gene <- rownames(vsd_assay)
  
  ## Merge with normalized count data
  resdata <- merge(as.data.frame(res), vsd_assay, by="row.names", sort=FALSE)
  names(resdata)[1] <- "Gene"
  rownames(resdata) <- resdata$Gene
  resdata <- resdata[-1]
  
  #Merge with annotation (genesymbol and geneType)
  resdata_annot <- merge(as.data.frame(resdata),as.data.frame(annotdata),by="row.names",all.x= TRUE)
  rownames(resdata_annot) <- resdata_annot$Row.names
  resdata_annot <- resdata_annot[-1]
  
  # Selection on padj <= 0.05
  res$padj <- ifelse(is.na(res$padj), 1, res$padj)
  sigGenes <- rownames(res[res$padj <= .05,])
  
  resdata_annot <- resdata_annot[resdata_annot$Gene %in% sigGenes,]
  
  mat = resdata_annot[ head(order(resdata_annot$padj),50), ] # select the top 50 genes with the lowest padj
  rownames(mat) <- paste(mat$Gene, mat$name, sep="_")
  mat$log2FoldChange <- NULL
  mat$lfcSE <- NULL
  mat$stat <- NULL
  mat$pvalue <- NULL
  mat$baseMean <- NULL
  mat$padj <- NULL
  mat$Gene <- NULL
  mat$name <- NULL
  mat$type <- NULL
  mat$description <- NULL
  
  mat = mat - rowMeans(mat) # Subtract the row means from each value
  # Optional, but to make the plot nicer:
  df = as.data.frame(colData(vsd)[,c("Group")]) # Create a dataframe with a column of the conditions
  colnames(df) = "Group" # Rename the column header
  rownames(df) = colnames(mat) # add rownames
  # and plot the actual heatmap
  png(paste("rnaseq_ferroptomics/DifferentialAnalysis/genes_heatmap_",celline,"_",contrast,"_", affix, ".png", sep=""), 1200, 1000, pointsize=20)
  pheatmap(mat, annotation_col=df)
  dev.off()
}
```
```{r}

for (i in seq_along(dds_deseq_list)) {
  dds <- dds_deseq_list[[i]]
  res <- res_RvsS_lfc_list[[i]]
  affix <- affixes[i]
  print(affix)
  
  rld <- vst(dds)
  
  # make the lists
  upgenes <- rownames(head(res[ order( res$log2FoldChange ), ], n=20))
  downgenes <- rownames(head(res[ order( -res$log2FoldChange ), ], n=20))

  # this gives us the rows we want
  rows <- match(upgenes, row.names(rld))
  mat <- assay(rld)[rows,]
  mat <- mat - rowMeans(mat)

  # the labels are hard coded at the moment :(
  df <- as.data.frame(colData(rld)[c(condition_name)])
  png(paste("rnaseq_ferroptomics/DifferentialAnalysis/top20UP_",celline,"_",contrast,"_", affix, ".png", sep=""))
  pheatmap(mat, fontsize=5, annotation_col=df, main=paste(affix,"top 20 up genes"))
  dev.off()

  # this gives us the rows we want
  rows <- match(downgenes, row.names(rld))
  mat <- assay(rld)[rows,]
  mat <- mat - rowMeans(mat)

  df <- as.data.frame(colData(rld)[c(condition_name)])
  png(paste("rnaseq_ferroptomics/DifferentialAnalysis/top20DOWN_",celline,"_",contrast,"_", affix, ".png", sep=""))
  pheatmap(mat, fontsize=5, annotation_col=df, main=paste(affix,"top 20 down genes"))
  dev.off()
}
```

# Kegg analysis

```{r setup_go_kegg, echo=FALSE, message=FALSE, warning=FALSE}
#Gene set analysis vs. gene set enrichment analysis

#The two predominantly used enrichment methods are:

#Overrepresentation analysis (ORA), testing whether a gene set contains disproportional many genes of significant expression change, based on the procedure outlined in the first section
#Gene set enrichment analysis (GSEA), testing whether genes of a gene set accumulate at the top or bottom of the full gene vector ordered by direction and magnitude of expression change Subramanian et al., 2005

dir.create("rnaseq_ferroptomics/KEGG")

search_kegg_organism(str = "human", by= "common_name")

ah <- AnnotationHub()
ahDb <- query(ah, pattern = c("homo sapiens", "EnsDb", 	105))
relevant_idx <- which(base::grepl(pattern = 105, x = ahDb$title ))
hsa_db <- ahDb[[relevant_idx]]
rm(ahDb); rm(ah)

# Get kegg pathways
hsa_gs <- getGenesets(org = "hsa", db = "kegg", cache = FALSE, go.mode = "biomart") 
# get GO annotation
hsa_go <-getGenesets(org = "hsa", db = "go", cache = FALSE) 


resDE_GE <- as.data.frame(res)
resDE_GE$gene_id <- rownames(res)
resDE_GE  <- resDE_GE %>% 
             mutate( abs_log2FoldChange =  abs(resDE_GE$log2FoldChange)) %>%               
             arrange(desc(abs_log2FoldChange)) %>% 
             dplyr::filter(padj <= 10^-2)

resDE_GE <- resDE_GE[complete.cases(resDE_GE[ , 5:6]),]
resDE_GE$entrez = mapIds(hsa_db,
                      keys=as.character(resDE_GE$gene_id),
                      column="ENTREZID",
                      keytype="GENEID",
                      multiVals="first")

resDE_GE$symbol = mapIds(hsa_db,
                      keys=as.character(resDE_GE$gene_id),
                      column="SYMBOL",
                      keytype="GENEID",
                      multiVals="first")

#Necessary for the GAGE run input variables
foldchanges = resDE_GE$log2FoldChange
names(foldchanges) = resDE_GE$entrez

write.csv(resDE_GE, file="rnaseq_ferroptomics/KEGG/DE_GE.csv")
```


```{r}
keggs_same <- gage(exprs = foldchanges, gsets = hsa_gs, ref = NULL, samp = NULL, same.dir = FALSE)
print('Pathways analysis result')
print('Top 10 kegg pathway')
print(head(data.frame(keggs_same$greater,10)))
```


```{r}
print('Top 10 GO term')
go_same <- gage(exprs = foldchanges, gsets = hsa_go, ref = NULL, samp = NULL, same.dir = FALSE)
print(head(data.frame(go_same$greater,10)))
```

# fsgea analysis (GSEA for pathway and Gene Ontology)

```{r}

#Using fgsea toolkit

#Get data from https://data.broadinstitute.org/gsea-msigdb/msigdb/release/7.1/
#INFO on https://stephenturner.github.io/deseq-to-fgsea/
# http://bioconductor.org/packages/release/bioc/html/fgsea.html
# and https://www.gsea-msigdb.org/gsea/msigdb/collections.jsp#C2
# https://www.biorxiv.org/content/10.1101/060012v1

#Gene set analysis vs. gene set enrichment analysis
#The two predominantly used enrichment methods are:

#Overrepresentation analysis (ORA), testing whether a gene set contains disproportional many genes of significant expression change, based on the procedure outlined in the first section
#Gene set enrichment analysis (GSEA), testing whether genes of a gene set accumulate at the top or bottom of the full gene vector ordered by direction and magnitude of expression change Subramanian et al., 2005


######Set specific result (see table above) to res and process

# Get extra ID information => Gene SYMBOL

dir.create("rnaseq_ferroptomics/fsgea")

search_kegg_organism(str = "human", by= "common_name")

ah <- AnnotationHub()
ahDb <- query(ah, pattern = c("homo sapiens", "EnsDb", 	105))     
relevant_idx <- which(base::grepl(pattern = 105, x = ahDb$title ))
hsa_db <- ahDb[[relevant_idx]]
rm(ahDb); rm(ah)


resDE_GE <- as.data.frame(res)
resDE_GE$gene_id <- rownames(res)

write.csv(resDE_GE, file="rnaseq_ferroptomics/DifferentialAnalysis/DE_GE.csv")

#Add the gene symbol annotation
resDE_GE$symbol = mapIds(hsa_db,
                      keys=as.character(resDE_GE$gene_id),
                      column="SYMBOL",
                      keytype="GENEID",
                      multiVals="first")

# Only top 2000 expressed genes are taken into account (otherwise the GSEA ranking has too many elements with equal stats (because of low expression), messing up the ranking))
res2 <- resDE_GE %>% 
  top_n(2000, baseMean) %>%
  dplyr::select(symbol, baseMean) %>%
  na.omit() %>% 
  distinct() %>% 
  group_by(symbol) %>% 
  summarize(stat=mean(baseMean))
res2

ranks <- deframe(res2)
head(ranks, 20)

### PATHWAY ANALYSIS - KEGG
fgseaRes <- fgsea(pathways=gmtPathways("rnaseq_ferroptomics/fsgea/c2.cp.kegg.v7.1.symbols.gmt"), ranks) %>% 
  as_tibble() %>% 
  arrange(padj)

fgseaResTidy <- fgseaRes %>%
  as_tibble() %>%
  dplyr::filter(padj <= 0.25) %>%
  arrange(desc(abs(NES)))

fwrite(fgseaResTidy, file=paste("rnaseq_ferroptomics/fsgea/GSEA_PATHWAY_KEGG_", celline,"_", contrast, ".csv", sep="" ), sep=",", sep2=c("", " ", ""))


### PATHWAY ANALYSIS
fgseaRes <- fgsea(pathways=gmtPathways("rnaseq_ferroptomics/fsgea/c2.cp.v7.1.symbols.gmt"), ranks) %>% 
  as_tibble() %>% 
  arrange(padj)
  
fgseaResTidy <- fgseaRes %>%
  as_tibble() %>%
  dplyr::filter(padj <= 0.25) %>%
  arrange(desc(abs(NES)))

fwrite(fgseaResTidy, file=paste("rnaseq_ferroptomics/fsgea/GSEA_PATHWAY_", celline,"_", contrast, ".csv", sep="" ), sep=",", sep2=c("", " ", ""))

 
### GO ANALYSIS
#BIOLOGICAL PROCESS
fgseaRes <- fgsea(pathways=gmtPathways("rnaseq_ferroptomics/fsgea/c5.bp.v7.1.symbols.gmt"), ranks) 

fgseaResTidy <- fgseaRes %>%
  as_tibble() %>%
  dplyr::filter(padj <= 0.25) %>%
  arrange(desc(abs(NES)))

fwrite(fgseaResTidy, file=paste("rnaseq_ferroptomics/fsgea/GSEA_GO_BP_", celline,"_", contrast, ".csv", sep="" ), sep=",", sep2=c("", " ", ""))
 #write.csv(fgseaResTidySel, file=paste("GSEA_GO_BP_", celline,"_", contrast, ".csv", sep="" ))

 #MOLECULAR FUNCTION
 fgseaRes <- fgsea(pathways=gmtPathways("rnaseq_ferroptomics/fsgea/c5.mf.v7.1.symbols.gmt"), ranks) 

fgseaResTidy <- fgseaRes %>%
  as_tibble() %>%
  dplyr::filter(padj <= 0.25) %>%
  arrange(desc(abs(NES)))
fwrite(fgseaResTidy, file=paste("rnaseq_ferroptomics/fsgea/GSEA_GO_MF_", celline,"_", contrast, ".csv", sep="" ), sep=",", sep2=c("", " ", ""))
 
#CELLULAR COMPONENT
 fgseaRes <- fgsea(pathways=gmtPathways("rnaseq_ferroptomics/fsgea/c5.cc.v7.1.symbols.gmt"), ranks) 

fgseaResTidy <- fgseaRes %>%
  as_tibble() %>%
  dplyr::filter(padj <= 0.25) %>%
  arrange(desc(abs(NES)))


fwrite(fgseaResTidy, file=paste("rnaseq_ferroptomics/fsgea/GSEA_GO_CC_", celline,"_", contrast, ".csv", sep="" ), sep=",", sep2=c("", " ", ""))
```

# WebGestalt  (ORA, Over-representation Analysis)

```{r}
#Check available genesets
listGeneSet("hsapiens")

dir <- "rnaseq_ferroptomics/WebGestalt_ORA/"
dir.create(dir)

############################################
#Get the diff_expression results in genelist
############################################

#Prepare Annotation hub
ah <- AnnotationHub()
ahDb <- query(ah, pattern = c("Homo sapiens", "EnsDb", 	105))     
relevant_idx <- which(base::grepl(pattern = 105, x = ahDb$title ))
hsa_db <- ahDb[[relevant_idx]]
rm(ahDb); rm(ah)

#Filter DE results based on padj and abs(lo2FoldChange)
resDE_GE <- as.data.frame(res)
resDE_GE$gene_id <- rownames(res)
resDE_GE  <- resDE_GE %>% 
             mutate( abs_log2foldChange =  abs(resDE_GE$log2FoldChange)) %>%               
             arrange(desc(abs_log2foldChange)) %>% 
             dplyr::filter(padj <= 0.05)

#Fetch Entrez ID
resDE_GE <- resDE_GE[complete.cases(resDE_GE[ , 5:6]),]
resDE_GE$symbol = mapIds(hsa_db,
                  keys=as.character(resDE_GE$gene_id),
                  column="SYMBOL",
                  keytype="GENEID",
                  multiVals="first")

resDE_symbol <- resDE_GE%>%
            dplyr::select(symbol)
resDE_gene_id <- resDE_GE%>%
            dplyr::select(gene_id)

GeneListSymbol <- paste("ORA_symbol_list_",contrast,".txt",sep="")
GeneListGeneID <- paste("ORA_geneid_list_",contrast,".txt",sep="")
GeneRefListSymbol <- paste(dir,"ORA_symbol_reflist_",contrast,".txt",sep="")
GeneRefListGeneID <- paste(dir,"ORA_geneid_reflist_",contrast,".txt",sep="")
  
RankFileName <- paste("GSEA_symbol_",contrast,".rnk",sep="")

write.table(resDE_symbol, file=GeneListSymbol, quote=FALSE, row.names=FALSE, col.names=FALSE)
write.table(resDE_gene_id, file=GeneListGeneID, quote=FALSE, row.names=FALSE, col.names=FALSE)


#Obtain the reference expression set
resDE_GE <- as.data.frame(res)
resDE_GE$gene_id <- rownames(res)

#Fetch Entrez ID
resDE_GE <- resDE_GE[complete.cases(resDE_GE[ , 5:6]),]
resDE_GE$symbol = mapIds(hsa_db,
                  keys=as.character(resDE_GE$gene_id),
                  column="SYMBOL",
                  keytype="GENEID",
                  multiVals="first")
resDE_symbol <- resDE_GE%>%
            dplyr::select(symbol)
resDE_gene_id <- resDE_GE%>%
            dplyr::select(gene_id)

write.table(resDE_symbol, file=GeneRefListSymbol, quote=FALSE, row.names=FALSE, col.names=FALSE)
write.table(resDE_gene_id, file=GeneRefListGeneID, quote=FALSE, row.names=FALSE, col.names=FALSE)

#Use View(listGeneSet()) to check on the different functional categories
#pathway_KEGG
#geneontology_Molecular_Function_noRedundant
#geneontology_Biological_Process_noRedundant

#####
#ORA#
#####

enrichResult <- WebGestaltR(enrichMethod="ORA", organism="hsapiens",
enrichDatabase="pathway_KEGG", interestGeneFile=GeneListSymbol,
interestGeneType="genesymbol", referenceGeneFile=GeneRefListSymbol,
referenceGeneType="genesymbol", isOutput=TRUE, sigMethod="top", minNum=5, reportNum = 40, outputDirectory=dir, projectName=paste("ORA_",contrast,"_pathway_kegg",sep=""))

enrichResult <- WebGestaltR(enrichMethod="ORA", organism="hsapiens",
enrichDatabase="geneontology_Molecular_Function_noRedundant", interestGeneFile=GeneListSymbol,
interestGeneType="genesymbol", referenceGeneFile=GeneRefListSymbol,
referenceGeneType="genesymbol", isOutput=TRUE, sigMethod="top", minNum=5, reportNum = 40, outputDirectory=dir, projectName=paste("ORA_",contrast,"_GO_MF",sep=""))

enrichResult <- WebGestaltR(enrichMethod="ORA", organism="hsapiens",
enrichDatabase="geneontology_Biological_Process_noRedundant", interestGeneFile=GeneListSymbol,
interestGeneType="genesymbol", referenceGeneFile=GeneRefListSymbol,
referenceGeneType="genesymbol", isOutput=TRUE, sigMethod="top", minNum=5, reportNum = 40, outputDirectory=dir, projectName=paste("ORA_",contrast,"_GO_BP",sep=""))

enrichResult <- WebGestaltR(enrichMethod="ORA", organism="hsapiens",
enrichDatabase="geneontology_Cellular_Component_noRedundant", interestGeneFile=GeneListSymbol,
interestGeneType="genesymbol", referenceGeneFile=GeneRefListSymbol,
referenceGeneType="genesymbol", isOutput=TRUE, sigMethod="top", minNum=5, reportNum = 40, outputDirectory=dir, projectName=paste("ORA_",contrast,"_GO_CP",sep=""))

```

# WebGestalt  (GSEA, Gene Set Enrichment Analysis)

```{r}

dir <- "rnaseq_ferroptomics/WebGestalt_GSA/"
dir.create(dir)

#The count matrix
counts_dds <-as.data.frame(counts(dds, normalized=TRUE))

#Prepare Annotation hub
ah <- AnnotationHub()
ahDb <- query(ah, pattern = c("homo sapiens", "EnsDb", 	105))     
relevant_idx <- which(base::grepl(pattern = 105, x = ahDb$title ))
hsa_db <- ahDb[[relevant_idx]]
rm(ahDb); rm(ah)

#Obtain the reference expression set
resDE_GE <- as.data.frame(res)
resDE_GE$gene_id <- rownames(res)
counts_dds$gene_id <-rownames(counts_dds)

#Fetch Entrez ID
resDE_GE <- resDE_GE[complete.cases(resDE_GE[ , 5:6]),]
resDE_GE$symbol = mapIds(hsa_db,
                  keys=as.character(resDE_GE$gene_id),
                  column="SYMBOL",
                  keytype="GENEID",
                  multiVals="first")

counts_dds$symbol = mapIds(hsa_db,
                  keys=as.character(counts_dds$gene_id),
                  column="SYMBOL",
                  keytype="GENEID",
                  multiVals="first")
counts_dds_GSEA <- cbind(counts_dds$symbol,counts_dds$gene_id,counts_dds) ##SV: changed 15 to 6 here, 6 samples considered
colnames(counts_dds_GSEA)[colnames(counts_dds_GSEA) == "counts_dds$symbol"] <- "name"
colnames(counts_dds_GSEA)[colnames(counts_dds_GSEA) == "counts_dds$gene_id"] <- "description"

CountsFileName <- paste(dir, "GSEA_symbol_",contrast,".txt",sep="")
write.table(counts_dds_GSEA, file=CountsFileName, sep = "\t", quote=FALSE, row.names=FALSE, col.names=TRUE)


# Only top 1000 expressed genes are taken into account (otherwise the GSEA ranking has too many elements with equal stats (because of low expression), messing up the ranking))
res2 <- resDE_GE %>% 
  #top_n(1000, baseMean) %>%
  dplyr::select(symbol, baseMean) %>%
  na.omit() %>% 
  distinct() %>% 
  group_by(symbol) %>% 
  #summarize(stat=mean(stat))  %>%
  arrange(desc(baseMean))
res2

RankFileName <- paste(dir,"GSEA_symbol_",contrast,".rnk",sep="")
write.table(res2, file=RankFileName, sep = "\t", quote=FALSE, row.names=FALSE, col.names=FALSE)

######
#GSEA#
######

###Pathway

#KEGG
enrichResult <- WebGestaltR(enrichMethod="GSEA", organism="hsapiens",
  enrichDatabase="pathway_KEGG", interestGeneFile=RankFileName, dagColor = "continuous",
  interestGeneType="genesymbol", sigMethod="fdr", fdrThr=0.1, minNum=5, reportNum = 40,
  outputDirectory=dir, projectName = paste("GSEA_",contrast,"_Pathway_KEGG",sep=""))

#WikiPathway
enrichResult <- WebGestaltR(enrichMethod="GSEA", organism="hsapiens",
  enrichDatabase="pathway_Wikipathway", interestGeneFile=RankFileName, dagColor = "continuous",
  interestGeneType="genesymbol", sigMethod="fdr", fdrThr=0.1, minNum=5, reportNum = 40,
  outputDirectory=dir, projectName = paste("GSEA_",contrast,"_Pathway_Wiki",sep=""))

#Panther
# enrichResult <- WebGestaltR(enrichMethod="GSEA", organism="mmusculus",
#   enrichDatabase="pathway_Panther", interestGeneFile=RankFileName, dagColor = "continuous",
#   interestGeneType="genesymbol", sigMethod="fdr", fdrThr=0.1, minNum=3, reportNum = 40,
#   outputDirectory=dir, projectName = paste("GSEA_",contrast,"_Pathway_Panther",sep=""))

#Reactome
enrichResult <- WebGestaltR(enrichMethod="GSEA", organism="hsapiens",
  enrichDatabase="pathway_Reactome", interestGeneFile=RankFileName, dagColor = "continuous",
  interestGeneType="genesymbol", sigMethod="fdr", fdrThr=0.1, minNum=5, reportNum = 40,
  outputDirectory=dir, projectName = paste("GSEA_",contrast,"_Pathway_Reactome",sep=""))


###GO

#Molecular Function
enrichResult <- WebGestaltR(enrichMethod="GSEA", organism="hsapiens",
  enrichDatabase="geneontology_Molecular_Function_noRedundant", interestGeneFile=RankFileName, dagColor = "continuous",
  interestGeneType="genesymbol", sigMethod="fdr", fdrThr=0.1, minNum=5, reportNum = 40,
  outputDirectory=dir, projectName = paste("GSEA_",contrast,"_GO_MF",sep=""))

#Biological process
enrichResult <- WebGestaltR(enrichMethod="GSEA", organism="hsapiens",
  enrichDatabase="geneontology_Biological_Process_noRedundant", interestGeneFile=RankFileName, dagColor = "continuous",
  interestGeneType="genesymbol", sigMethod="fdr", fdrThr=0.1, minNum=5, reportNum = 40,
  outputDirectory=dir, projectName = paste("GSEA_",contrast,"_GO_BP",sep=""))

#Cellular component
enrichResult <- WebGestaltR(enrichMethod="GSEA", organism="hsapiens",
  enrichDatabase="geneontology_Cellular_Component_noRedundant", interestGeneFile=RankFileName, dagColor = "continuous",
  interestGeneType="genesymbol", sigMethod="fdr", fdrThr=0.1, minNum=5, reportNum = 40,
  outputDirectory=dir, projectName = paste("GSEA_",contrast,"_GO_CC",sep=""))

```